Parametricity Consequences of GADTs Representations

### 2021-03-01 20-gadts/GADT-story.tex
small revisions

A semanticist, on the other hand, is likely to follow~\cite{bfss90}
and expect GADTs to behave as their names suggests --- i.e., as
generalizations of ADTs. The semanticist is thus likely to expect a
GADT to be able to be used as a ADT is used --- i.e., as a shape
filled with data that can be changed without also changing the
shape. To be used in this way, a GADT must be viewed in terms of its
primitive representation. In fact, the functorial completion of a GADT
inherent in its primitive representation is exactly what is needed to
ensure that the GADT supports a $\mathsf{map}$ function, i.e., is
interpreted as a true categorical functor, and can therefore be made
an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
however, that since the functorial completion of a GADT includes only
those data elements that are derivable via repeated applications of
$\mathsf{map}$ starting from those specified by its syntax, primitive
GADTs need not, in general, be nested types. For example, the
functorial completion of the GADT $\mathsf{G}$ defined by
\begin{equation}\label{eq:G}
\begin{array}{l}
\mathsf{data\,G\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c : G\,1}
\end{array}
\end{equation}
{\color{blue} Use $\top$ instead of $1$ throughout?} includes elements
at any instance $\mathsf{G\,a}$ for any type $\mathsf{a} \not =
\mathsf{0}$ since there is always a function from $\mathsf{1}$ to such
an $\mathsf{a}$, but does not include any terms at instance
$\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
syntactic specifiation of $\mathsf{G}$, and is not inhabited via
functorial completion because it is not possible to define a function
from $\mathsf{1}$ into $\mathsf{0}$. {\color{blue} Use $\bot$ instead
  of $0$ here?}

@@

Importantly, the function $\mathit{Equal}\,a$ cannot be made into a
functor. Equivalently, $\mathsf{Equal\,a}$ (in either representation)
cannot be made an instance of Haskell's $\mathsf{Functor}$
class. Indeed, if $\mathsf{Equal\,a}$ supported a function
\[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
  Equal\,a\,c}\]
then defining
\[\begin{array}{l}
\mathsf{eqElim :: Equal\, a\, b \to b \to a}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct an element 
\[\begin{array}{l}
\mathsf{emptyFromMapEquala :: (\forall b\, c.\, (b \to c) \to Equal\, a\,
  b \to Equal\, a \,c) \to Void}\\ 
\mathsf{emptyFromMapEquala \,map_{Equal\,a} = eqElim\, (map_{Equal\,a}\,
  absurd\, Refl)\, ()}  
\end{array}\]
of type $\mathsf{Void}$. {\color{blue} Use same type conventions as in
  Intro.} But this should be impossible. It is immediate that a GADT
that is not actually a nested type cannot support a $\mathsf{map}$
function (or a rearrange-transform property) either. In the case of
$\mathsf{Seq}$, this is already intuitively evident from its Haskell
definition as discussed in Section~\ref{sec:intro}.

@@

Consider, again, the GADT $\mathsf{Seq}$. Because its type argument
varies in the instances of $\mathsf{Seq}$ appearing in the types of
its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be
modelled as the fixpoint of any first-order functor. As shown
in~\cite{jp19}, it can, however, be modelled as a solution to the
higher-order fixpoint equation
\[H\,f\,a = a \,+\,(Lan_{\lambda c d. c \times d}\, \lambda
c d. f c \times f d)\,a\]
{\color{red} Check this equation against next displayed one below.}
where $Lan_K \,F$ is the left Kan extension of
the functor $F$ along the functor $K$. In general, the left Kan
extension $Lan_K \,F : {\cal E} \to {\cal D}$ of $F : {\cal C} \to
{\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best functorial
approximation to $F$ that factors through $K$. Here, ``best functorial
approximation'' means that $Lan_K \,F$ is the smallest functor that
both extends the image of $K$ to $\cal D$ and coincides with $F$ on
$\cal C$, in the sense that, for any other such functor $H$, there is
a morphism of functors (i.e., a natural transformation) from $Lan_K
\,F$ to $H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $F : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $F$ along $K$} is a
functor $\mathit{Lan}_K\,F : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : F \to (\mathit{Lan}_K\,F) \circ K$ such
that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : F \to H \circ K$, there exists a unique
natural transformation $\delta : \mathit{Lan}_K\,F \to H$ such that
$(\delta K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{F}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,F}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
\end{tikzcd}\]
\end{definition}

To represent GADTs as primitive data types in a setting in which types
are interpreted as sets, a calculus must support a primitive construct
$\mathsf{Lan}$ such that the type constructor $\mathsf{Lan_k\,f}$ is
the syntactic reflection of the left Kan extension $\Lan_K \,F$ of the
functor $F$ interpreting $\mathsf{f}$ along the functor $K$
interpreting $\mathsf{k}$. In this setting, the categories
$\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be
$\mathit{Set}$. Using $\mathsf{Lan}$ we can then rewrite the type of a
constructor $\mathsf{c :: f a \to g(k\,a)}$ as $\mathsf{c ::
  (Lan_k\,f)\,a \to g\,a}$ since morphisms (i.e., natural
transformations) from $F$ to $G \circ K$ are in one-to-one
corresondence with those from $\mathit{Lan}_K\,F$ to $G$ (see,
e.g.,~\cite{rie16}). That is, writing $F \Rightarrow G$ for the set of
natural transformations from a functor $F$ to a functor $G$, we have
\begin{equation}\label{eq:nat-transfs}
F \Rightarrow G \circ K\; \simeq \; Lan_K\,F \Rightarrow G
\end{equation}
The calculus must also support a primitive type
constructor $\mathsf{\mu}$ that is the syntactic reflection of the
(now higher-order) fixpoint operator on
$\mathit{Set}^\mathit{Set}$. Using $\mu$ and $\Lan$ we can then
represent a GADT as a higher-order higher-order fixpoint. For example,
we can represent the GADT $\mathsf{Seq}$ as
\[\mathsf{Seq\,a} = \mathsf{(\mu \phi.\lambda b.\, b + (Lan_{\lambda c
    d. c \times d} \lambda c d. \phi c \times \phi d)\,b)\,a}\] The
fact that $Lan_K\,G$ is the best functorial approximation to $G$
factoring through $K$ means that the type constructor
$\mathsf{Lan_k\,g}$ computes the smallest collection of data that is
generated by the corresponding GADT data constructor's syntax and also
supports a $\mathsf{map}$ function. The primitive representation of
any GADT thus comprises the smallest data type that both includes the
data specified by that GADT's syntax and also supports a
$\mathsf{map}$ function (and thus a rearrange-transform property). In
the primitive view, then, GADTs are generally underspecified by their
syntax.

@@

\begin{example}\label{ex:g1}
Syntactically, the GADT $\mathsf{G}$ defined in~\eqref{eq:G} comprises
a single data element, namely $\mathsf{c :: G \, 1}$. This is captured
by $\mathsf{G}$'s Church encoding
\[\mathsf{G\,a} = \mathsf{Equal\,1\,a}\]
from~\cite{atk12}. As its Church encoding makes clear, $\mathsf{G}$'s
effect is simply to test its argument for equality against the unit
type $\mathsf{1}$. {\color{blue} Use same notation for unit type as in
  intro.} The interpretation of the Church encoding of $\mathsf{G}$
is thus, in accordance with Section~\ref{sec:CEs}, the function whose
value is a singleton set when the interpretation of $\mathsf{a}$ is
$1$ and the empty set otherwise. This perfectly accords with
$\mathsf{G}$'s syntax, which indeed delivers just one data element
when $\mathsf{a}$ is $\mathsf{1}$ and no elements otherwise.

@@

For primitive representations of GADTs the story is completely
different. First off, we have that {\color{blue} the interp of Lan
  must be full Lan}. Indeed,... {\color{blue} MORE!!}
\begin{proposition}\label{prop:full-lans}
Full Lan prop goes here. Explain cutting down.
\end{proposition}
\noindent
In particular, both the set interpretation and the relational
interpetation of a GADT must involve the entireties of any Lans in
their fixpoint representations. We therefore have:

@@

Consider the relation $R = (1, 2, 1 \times 2)$. We expect
$(\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
*\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain {\color{blue}
  $1$}. Since left Kan extensions preserve projections~\cite{rie16},
we can compute the domain as
\[\pi_1\,\big( (\mathit{\Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
*\!.\,\mathit{Eq}_1) \, R \big)\;=\; (\mathsf{\Lan}_{\lambda
  *\!.\,\mathit{Eq}_1} \lambda *\!. \,1)\, 1\]

(notice that the left
Kan extension of $1 : \ast \to \set$ along $\Eq_1 : \ast \to \rel$ is
a functor $\rel \to \set$).

Consider the relation $R = (1, \emptyset, \emptyset)$.  We would thus
expect $(\mathsf{\Lan}_{\Eq_1} \Eq_1) R$ to be a relation with domain
$1$.  Because the left Kan extension preserves
projections~\cite{rie16} we can compute the domain as
$(\mathsf{\Lan}_{\Eq_1} 1) R$ (notice that the left Kan extension of
$1 : \ast \to \set$ along $\Eq_1 : \ast \to \rel$ is a functor $\rel
\to \set$).

>>>>

A semanticist, on the other hand, is likely to follow~\cite{bfss90}
and expect GADTs to behave as their names suggests --- i.e., as
generalizations of ADTs. The semanticist is thus likely to expect a
GADT to be able to be used as a ADT is used --- i.e., as a shape
filled with data that can be changed without also changing the
shape. To be used in this way, a GADT must be viewed in terms of its
primitive representation. In fact, the functorial completion of a GADT
inherent in its primitive representation is exactly what is needed to
ensure that the GADT supports a $\mathsf{map}$ function, i.e., is
interpreted as a true categorical functor, and can therefore be made
an instance of Haskell's $\mathsf{Functor}$ class. It is worth noting,
however, that since the functorial completion of a GADT includes only
those data elements that are derivable via repeated applications of
$\mathsf{map}$ starting from those specified by its syntax, primitive
GADTs need not, in general, be nested types. For example, the
functorial completion of the GADT $\mathsf{G}$ defined by
\begin{equation}\label{eq:G}
\begin{array}{l}
\mathsf{data\,G\,a\,where}\\
\mathsf{\;\;\;\;\;\;\;\;c :: G\,1}
\end{array}
\end{equation}
Here, $\mathsf{1}$ is the unit type, whose single element is also
denoted by $\mathsf{1}$, includes elements at any instance
$\mathsf{G\,a}$ for any type $\mathsf{a}$ that is not the empty type
$\mathsf{0}$ since there is always a function from $\mathsf{1}$ to
such an $\mathsf{a}$, but does not include any terms at instance
$\mathsf{G\,0}$. Indeed, $\mathsf{G\,0}$ is not inhabited via the
syntactic specifiation of $\mathsf{G}$, and is not inhabited via
functorial completion because it is not possible to define a function
from $\mathsf{1}$ into $\mathsf{0}$.

@@

Importantly, the function $\mathit{Equal}\,a$ cannot be made into a
functor. Equivalently, $\mathsf{Equal\,a}$ (in either representation)
cannot be made an instance of Haskell's $\mathsf{Functor}$
class. Indeed, if $\mathsf{Equal\,a}$ supported a function
\[\mathsf{map_{Equal\,a} :: (b \to c) \to Equal\,a\,b \to
  Equal\,a\,c}\]
then defining
\[\begin{array}{l}
\mathsf{eqElim :: Equal\, a\, b \to b \to a}\\
\mathsf{eqElim\, Refl\, x = x} 
\end{array}\]
would allow us to construct
an element
\[\mathsf{eqElim\, (map_{Equal\,Void}\, absurd\, Refl)\, 1}\]
of type $\mathsf{Void}$, where $\mathsf{absurd :: 0 \to c}$ is the
syntactic reflection of the empty function from the empty set into the
interpretation of $\mathsf{c}$. {\color{blue} Is there a way to say
  this without passing to semantics?} But this should be
impossible. It is immediate that a GADT that is not actually a nested
type cannot support a $\mathsf{map}$ function (or a
rearrange-transform property) either. In the case of $\mathsf{Seq}$,
this is already intuitively evident from its Haskell definition as
discussed in Section~\ref{sec:intro}.

@@

Consider, again, the GADT $\mathsf{Seq}$. Because its type argument
varies in the instances of $\mathsf{Seq}$ appearing in the types of
its data constructor $\mathsf{Pair}$, $\mathsf{Seq}$ cannot be
modelled as the fixpoint of any first-order functor. As shown
in~\cite{jp19}, it can, however, be modelled as a solution to the
higher-order fixpoint equation
\[H\,f\,a = a \,+\,(Lan_{\lambda c d. c \times d}\, \lambda
c d. f c \times f d)\,a\] {\color{red} Check this equation against
  next displayed one below.}  where $Lan_K \,F$ is the left Kan
extension of the functor $F$ along the functor $K$. In general, the
left Kan extension $Lan_K \,F : {\cal E} \to {\cal D}$ of $F : {\cal
  C} \to {\cal D}$ along $K : {\cal C} \to {\cal E}$ is the best
functorial approximation to $F$ that factors through $K$. Here, ``best
functorial approximation'' means that $Lan_K \,F$ is the smallest
functor that both extends the image of $K$ to $\cal D$ and agrees with
$F$ on $\cal C$, in the sense that, for any other such functor $H$,
there is a morphism of functors (i.e., a natural transformation) from
$Lan_K \,F$ to $H$. This is captured formally by the following
definition~\cite{mac71}:
\begin{definition}\label{def:lke}
If $F : {\cal C} \to {\cal D}$ and $K : {\cal C} \to {\cal E}$ are
functors, then the {\em left Kan extension of $F$ along $K$} is a
functor $\mathit{Lan}_K\,F : {\cal E} \to {\cal D}$ together with a
natural transformation $\eta : F \to (\mathit{Lan}_K\,F) \circ K$ such
that, for every functor $H : {\cal E} \to {\cal D}$ and natural
transformation $\gamma : F \to H \circ K$, there exists a unique
natural transformation $\delta : \mathit{Lan}_K\,F \to H$ such that
$(\delta K) \circ \eta = \gamma$. This is depicted in the diagram
\[\begin{tikzcd}[row sep = huge]
{\cal C}
\ar[rr, "{F}"{name=Fa, above}, ""{name=F, below}]
\ar[rd, "{K}"']
&& {\cal D} \\
& {\cal E}
\ar[Rightarrow, bend right = 25, from=F, "{\eta}"']
\ar[ur, bend left, "{\mathit{Lan}_{K}\,F}"{name=Lan, description}]
\ar[ur, bend right, "{H}"'{name=L, right}, ""'{name=Lr, right}]
\ar[Rightarrow, bend left = 45, from=Fa, to=Lr, "{\gamma}" near start]
\ar[Rightarrow, dashed, from=Lan, to=L, "{\delta}"']
\end{tikzcd}\]
\end{definition}

To represent GADTs as primitive data types in a setting in which types
are interpreted as sets, a calculus must support a primitive construct
$\mathsf{Lan}$ such that the type constructor $\mathsf{Lan_k\,f}$ is
the syntactic reflection of the left Kan extension $\Lan_K \,F$ of the
functor $F$ interpreting $\mathsf{f}$ along the functor $K$
interpreting $\mathsf{k}$. In this setting, the categories
$\mathcal{C}$, $\mathcal{D}$, and $\mathcal{E}$ must all be of theform $\mathit{Set}^n$ for some $n$. Using $\mathsf{Lan}$ we can then
rewrite the type of a constructor $\mathsf{c :: f a \to g(k\,a)}$ as
$\mathsf{c :: (Lan_k\,f)\,a \to g\,a}$ since morphisms (i.e., natural
transformations) from $F$ to $G \circ K$ are in one-to-one
corresondence with those from $\mathit{Lan}_K\,F$ to $G$ (see
e.g.,~\cite{rie16}). That is, writing $F \Rightarrow G$ for the set of
natural transformations from a functor $F$ to a functor $G$, we have
\begin{equation}\label{eq:nat-transfs}
F \Rightarrow G \circ K\; \simeq \; Lan_K\,F \Rightarrow G
\end{equation}
The calculus must also support a primitive type
constructor $\mathsf{\mu}$ that is the syntactic reflection of the
(now higher-order) fixpoint operator on
$\mathit{Set}^\mathit{Set}$. Using $\mu$ and $\Lan$ we can then
represent a GADT as a higher-order higher-order fixpoint. For example,
we can represent the GADT $\mathsf{Seq}$ as
\[\mathsf{Seq\,a} = \mathsf{(\mu \phi.\lambda b.\, b + (Lan_{\lambda c
    d. c \times d} \lambda c d. \phi c \times \phi d)\,b)\,a}\] The
fact that $Lan_K\,G$ is the best functorial approximation to $G$
factoring through $K$ means that the type constructor
$\mathsf{Lan_k\,g}$ computes the smallest collection of data that is
generated by the corresponding GADT data constructor's syntax and also
supports a $\mathsf{map}$ function. The primitive representation of
any GADT thus comprises the smallest data type that both includes the
data specified by that GADT's syntax and also supports a
$\mathsf{map}$ function (and thus a rearrange-transform property). In
the primitive view, then, GADTs are generally underspecified by their
syntax.

@@

\begin{example}\label{ex:g1}
Syntactically, the GADT $\mathsf{G}$ defined in~\eqref{eq:G} comprises
a single data element, namely $\mathsf{c :: G \, 1}$. This is captured
by $\mathsf{G}$'s Church encoding
\[\mathsf{G\,a} = \mathsf{Equal\,1\,a}\]
from~\cite{atk12}. As its Church encoding makes clear, $\mathsf{G}$'s
effect is simply to test its argument for equality against the unit
type $\mathsf{1}$. The interpretation of the Church encoding of
$\mathsf{G}$ is thus, in accordance with Section~\ref{sec:CEs}, the
function whose value is a singleton set when the interpretation of
$\mathsf{a}$ is $1$ and the empty set otherwise. This perfectly
accords with $\mathsf{G}$'s syntax, which indeed delivers just one
data element when $\mathsf{a}$ is $\mathsf{1}$ and no elements
otherwise.

@@

For primitive representations of GADTs the story is completely
different. First off, we have that {\color{blue} the interp of Lan
  must be full Lan}. Indeed,... {\color{blue} MORE!!}
\begin{proposition}\label{prop:full-lans}
Full Lan prop goes here. Explain cutting down. Or just argue intuitively?
\end{proposition}
\noindent
In particular, both the set interpretation and the relational
interpetation of a GADT must involve the entireties of any Lans in
their fixpoint representations. We therefore have:

@@

Consider the relation $R = (1, 2, 1 \times 2)$. We expect
$(\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
*\!.\,\mathit{Eq}_1)\, R$ to be a relation with domain $1$. Since left
Kan extensions preserve projections~\cite{rie16}, we can compute the
domain as
\[\pi_1\,\big( (\mathit{Lan}_{\lambda *\!.\,\mathit{Eq}_1} \lambda
*\!.\,\mathit{Eq}_1) \, R \big)\;=\; (\mathit{Lan}_{\lambda
  *\!.\,\mathit{Eq}_1} \lambda *\!. \,1)\, 1\] (Note that the left Kan
  extension of $\lambda *\!. \,1$ along $\lambda \!.\,\mathit{Eq}_1$ 
 is a functor from $\mathit{Rel}$ to $\mathit{Set}$.)

HERE!!!!!
